---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The data is loaded from the *"activity.csv"* file by using read.csv() into a data frame *aData*. There are 3 variables and 17568 observations. The variables are steps, date and interval. The total number of steps taken per day is calculated by preprocessing the data and stored in the variable *dailySteps*. The variable steps is summed for each date in the dataset. 
```{r}
options(scipen=999, digits=2)
aData<-read.csv("activity.csv")
dailySteps<-by(aData$steps, aData$date, sum)
```

## What is mean total number of steps taken per day?

The histogram of the total number of steps taken per day is shown below.
```{r dailysteps_hist, fig.height=4}
hist(dailySteps, breaks=50, main="Histogram of Total Steps taken per day", xlab="Total Steps per Day", col ="blue")
```
```{r mean_median}
mean_ds<-mean(dailySteps, na.rm=T)
median_ds<-median(dailySteps, na.rm=T)
```
The mean total number of steps taken per day is `r mean_ds` and the median value calculated is `r median_ds`.

## What is the average daily activity pattern?
The total number of steps taken for every 5 minute interval during the day is calculated in the variable *intervalSteps*. Here is the plot of the average number of steps taken for each 5 minute interval in a day, averaged across all days. Intervals with NA values were ignored in calculating the total steps during each interval.
```{r avg_steps, fig.height=4}
intervalSteps<-by(aData$steps, aData$interval, function(x) mean(x, na.rm=T))
plot(x=names(intervalSteps), y=intervalSteps, type='l', main="Daily Average Activity Pattern for Steps", xlab="5-min Intervals in a Day", ylab="Average Number of Steps")
```

## Imputing missing values
The number of missing values (NA) in the variable steps are calculated here.
```{r}
sum(is.na(aData$steps))
```
It was decided to impute the missing values for steps by replacing the value for a 5 min interval on a particular day with the average of the 5 min intervals across all days which is available from the variable *intervalSteps*. The index where *steps* is NA is obtained first. Then the *interval* value for where *steps* is NA is found and matched to the index of *intervalSteps*. This means we now have a valid total number of steps value for that interval where *steps* was NA. Next this value is averaged by total number of days, rounded and used to replace the original steps value of NA in a new data frame.
```{r imputeSteps}
nsteps<-aData$steps # Store steps as it is first into new variable
index<-is.na(aData$steps)==T # Index of all where steps is NA
#Find the index of corresponding interval when steps is NA in the intervalSteps list
#Now take that value from intervalSteps, divide by TotalDays and round it.
nsteps[index]<-round(intervalSteps[match(aData$interval[index],names(intervalSteps))] + 0.5)
# new data frame with new values of steps without any NA
nData <- data.frame(steps=nsteps, date=aData$date, interval=aData$interval)
```

Here is the new histogram of the total number of steps taken per day after the NA values in steps has been replaced by the new strategy of imputing data.
```{r new_hist, fig.height=4}
ndailySteps<-by(nData$steps, nData$date, sum)
hist(ndailySteps, breaks=50, main="Histogram of Total Steps taken per day (after imputing)", xlab="Total Steps per Day", col="red")
```
```{r new_mean}
mean_nds<-as.numeric(mean(ndailySteps, na.rm=T))
median_nds<-as.numeric(median(ndailySteps, na.rm=T))
```
The mean total number of steps taken per day is `r mean_nds` and the median value calculated is `r median_nds`.
Due to imputing missing values the mean value has remained almost the same but the median values of total number of steps per day was slightly higher.

## Are there differences in activity patterns between weekdays and weekends?
```{r weekday}
whichday<-weekdays(as.Date(nData$date)) #Find the day of the week from date
whichday[grep("Saturday|Sunday", whichday)]<-"weekend"
whichday[grep("Monday|Tuesday|Wednesday|Thursday|Friday", whichday)]<-"weekday"
nData<-cbind(nData,day=whichday) # Add new variable to the dataset
```
Here the average number of steps taken for each 5 minute interval for weekdays and weekends will be calculated using the aggregate() function. 
```{r panel}
nwkdays<-sum(nData$day=="weekday")/length(intervalSteps)
nwkends<-sum(nData$day=="weekend")/length(intervalSteps)
weekdata<-aggregate(steps ~ interval+day, mean, data=nData)
```
Here is the plot of the average steps taken for each 5 min interval during weekdays and weekends.
```{r plot}
library(ggplot2)
ggplot(weekdata, aes(y=steps, x=interval)) + geom_line(aes(col=day)) + 
  facet_grid(day~ .) + 
  ggtitle("Comparison of average steps taken between weekdays and weekends") + 
  xlab("5 min Intervals") + 
  ylab("Average number of steps taken")
```
